---
title: "A codebook for collecting data from archaeological site forms in Louisiana (v1.3)"
author: "Jonathan Paige^[New Mexico Consortium, jonathan.n.paige@Gmail.com]"
output:
  html_document:
    df_print: paged
  pdf_document:
    keep_tex: yes
    number_sections: yes
    toc: true
    toc_depth: 3
  word_document: default
header-includes:
- \usepackage{float}
- \floatplacement{figure}{hp}
- \usepackage{booktabs}
- \usepackage[font=small,labelfont=bf]{caption}
- \usepackage{setspace}\doublespacing
#abstract: |
keywords: |
  Louisiana archaeology; site forms; SHPO; codebook; text coding
highlights: |
  These are the highlights.
---

```{r, include=FALSE}
options(tinytex.verbose = TRUE)
```

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
```

# Version details:

This report was generated on `r Sys.time()`

The current Git commit details are:

```{r}
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())
```

# This codebook



This codebook was developed as part of a broader project investigating the reliability of large language models (LLMs) in performing text coding tasks in archaeology. It is targeted specifically to site forms on file with the Louisiana Division of Archaeology (Louisiana SHPO). Traits include presence/absence judgements about cultural components and artifact classes, as well as categorical and numeric extractions. These judgements must be based on a reading of the narrative and structured fields across site forms. This codebook provides guidance on making those judgements.

# Source Structure

Louisiana SHPO site forms contain a mix of structured fields (checkboxes, dropdown selections, coded entries) and free-text narrative sections. All are considered valid sources of evidence unless otherwise specified in individual trait definitions below. Pre-printed field labels on the form (e.g., column headers in artifact tallies) should not be treated as evidence that a trait is present; only filled-in values, checked boxes, or explicit narrative statements count as evidence. Furthermore, the site forms are appended to one another when new work is performed. In this context, we split the form into individual investigations, and the relevant pages reporting that investigation, and only provide those to the coder.The findings of an earlier investigation may find different things from a later investigation. So, each investigation should be coded independently. Prior to this coding process, we broke the pdfs up into just pages that are from a particular investigation, and which were likeliest to have relevant information about the traits being coded.


# Definitions

**Pre-Contact period**
Any occupation, use, or deposition of materials at a site that pre-dates sustained European contact in the relevant region of Louisiana. For most of Louisiana this is broadly taken as pre-dating the late 17th century (ca. pre-1700 CE), though recorders may use different cut-off dates. Both explicit statements and inferred dates (e.g., radiocarbon dates, diagnostic artifact types) are acceptable evidence.

**Post-Contact period**
Any occupation, use, or deposition of materials at a site that post-dates sustained European contact (broadly ca. post-1700 CE in Louisiana). Includes colonial French, Spanish, American, and later periods. Historic-era Euro-American, Creole, and African-descended communities may all fall within this period.

**Caddo period**
Occupation or material culture attributed to Caddo-speaking peoples or their archaeological antecedents in Louisiana, broadly spanning ca. 800–1700 CE (Late Woodland through Protohistoric), though boundaries are debated and vary by subregion. Includes both the Formative Caddo and Caddo tradition periods. Dating may be based on diagnostic ceramics, architecture, or radiocarbon dates.

**Archaic period**
Occupation or material culture attributed broadly to the Archaic, spanning roughly 10,000–3,000 BP in Louisiana, though exact boundaries vary by subregion and author. Includes Early, Middle, and Late Archaic subperiods. Dating may be based on diagnostic projectile point types, radiocarbon dates, or stratigraphic position.

**Chipped stone**
Any artifact produced through knapping (percussion or pressure flaking), including debitage (flakes, cores, shatter), tools (projectile points, scrapers, bifaces, etc.), and partially reduced nodules. Groundstone tools that also bear flaking scars do not automatically qualify; the primary reduction method should be knapping.

**Projectile point**
A bifacially or unifacially retouched tool interpreted as a tip or hafted element of a projectile weapon (spear, atlatl dart, or arrow). Formal typological assignments (e.g., "Gary point", "Perdiz point") are sufficient evidence. Fragments described as probable projectile points are acceptable. Drills and perforators should not be coded as projectile points unless the recorder explicitly identifies them as such.

**Fire-cracked rock (FCR)**
Thermally fractured rock resulting from exposure to heat, typically associated with cooking features, earth ovens, or hearths. May be described as burned rock, fire-cracked rock, FCR, heat-fractured rock, or similar terms. Rock that is merely discolored or scorched but not fractured does not qualify.

**NRHP eligibility**
Status of a site with regard to National Register of Historic Places inclusion. Eligible sites are eligible for inclusion; listed sites are listed; most sites may have undetermined or potential eligibility, barring more information typically gained through historical records or more intensive fieldwork focused on testing the site for research potential.



# Traits

## Pre-Contact component (Pre\_Conta)

**Data type:** binary

**Short Description:** Evidence of a pre-contact occupation or use of the site.

**Definition:** The site form records evidence either in structured fields or narrative text indicating that some portion of the site's archaeological record dates to the period before sustained European contact (broadly pre-1700 CE in Louisiana). This may include diagnostic artifacts, radiocarbon dates, or interpretive statements by the recorder.

**Inclusion criteria:** Evidence in text OR indicated in structured fields (e.g., period checkboxes, cultural affiliation fields, artifact type tallies referencing prehistoric types).

**Exclusion criteria:** Not described in text AND not indicated in structured fields. A site described as exclusively historic-era should be coded as absent.

**Typical exemplars:** Recorder checks "prehistoric" in the cultural period field. Narrative states cases like "the site contains a prehistoric component" or a cultural affiliation field lists "unknown prehistoric." Sites that are lithic scatters, or map onto any cultural phase before European contact.

**Atypical exemplars:** Historic sites with presence of prehistoric artifacts. A site is described as "multicomponent" and the prehistoric component is mentioned only in passing in the survey narrative.

**Close but no:** A historic-era site form that mentions "no prehistoric materials observed." Presence of burned rock or undated features alone, with no other indicator of pre-contact occupation. A site form for an extant historic structure that references "prehistoric regional context" in a background section without implying materials were recovered.

**Response format:** Binary. Return `true` if the trait is present, `false` if absent.



## Post-Contact component (Post\_Conta)

**Data type:** binary

**Short Description:** Evidence of a post-contact occupation or use of the site.

**Definition:** The site form records evidence indicating that some portion of the site's archaeological record dates to the period after sustained European contact (broadly post-1700 CE in Louisiana). This may include historic-era artifact types, documentary references, architectural remains, or explicit interpretive statements.

**Inclusion criteria:** Described in text OR indicated in structured fields (e.g., period checkboxes listing "historic," "colonial," "19th century," etc.; artifact tallies listing historic ceramics, glass, nails, or other temporally diagnostic historic materials).

**Exclusion criteria:** Not described in text AND not indicated in structured fields. A site described as exclusively prehistoric should be coded as absent.

**Typical exemplars:** Recorder checks "historic" in the period field. Narrative states "the site contains historic-era debris including European-made ceramics and cut nails." Cultural affiliation listed as "Anglo-American" or "French Colonial." A standing structure on the site dates to the 19th century.

**Atypical exemplars:** No explicit period designation, but the artifact inventory lists whiteware sherds, window glass, and wire nails with no mention of prehistoric materials. A site described as "multicomponent" where the historic component is noted only briefly.

**Close but no:** A prehistoric site whose background section discusses the broader historic settlement of the region. Presence of modern intrusions (e.g., fence posts, drainage pipes) noted but not interpreted as a historic-era occupation. A site where the recorder mentions a nearby historic structure but does not indicate association with the recorded site.

**Response format:** Binary. Return `true` if the trait is present, `false` if absent.



## Caddo period component (Caddo\_Unkn)

**Data type:** binary

**Short Description:** Evidence of a Caddo-period occupation, regardless of specific subperiod assignment.

**Definition:** The site form records evidence attributing some portion of the site's archaeological record to Caddo-speaking peoples or the Caddo archaeological tradition, broadly spanning ca. 800–1700 CE. The code name "Caddo_Unkn" reflects that the specific Caddo subperiod (e.g., Early, Middle, Late Caddo; Formative Caddo; Protohistoric) may be unknown or unspecified. This trait should be coded as present whenever Caddo affiliation is indicated, regardless of whether a finer subperiod is also assigned.

**Inclusion criteria:** Described in text OR indicated in structured fields (e.g., cultural affiliation field lists "Caddo," "Caddoan," or a named Caddo-tradition phase; artifact field lists Caddo ceramic types).

**Exclusion criteria:** Not described in text AND not indicated in structured fields. Archaic, Woodland, or other non-Caddo prehistoric components should not be coded as present here.

**Typical exemplars:** Cultural affiliation field states "Caddo." Narrative describes "Caddo ceramic sherds including engraved fine ware." A named phase such as "Haley," "Belcher," or "Kinsloe" is listed. The recorder assigns the site to the "Late Prehistoric" period and describes Caddo-tradition ceramic types.

**Atypical exemplars:** The recorder uses only the term "Late Prehistoric" without naming Caddo explicitly, but the artifact inventory includes ceramic types (e.g., Friendship Engraved, Hickory Fine Engraved) diagnostic of the Caddo tradition. A site recorded near known Caddo territory where the artifact assemblage is consistent with Caddo occupation but not explicitly labeled as such by the recorder.

**Close but no:** Sites described as Woodland-period only, even if geographically within the Caddo region. Sites with a general "prehistoric" designation and no specific Caddo indicators in either the structured fields or the narrative.

**Response format:** Binary. Return `true` if the trait is present, `false` if absent.



## Archaic period component (Arch\_Unkn)

**Data type:** binary

**Short Description:** Evidence of an Archaic-period occupation, regardless of specific subperiod (Early, Middle, or Late Archaic).

**Definition:** The site form records evidence attributing some portion of the site's archaeological record to the Archaic period, broadly spanning ca. 10,000–3,000 BP. The code name "Arch\_Unkn" reflects that the specific Archaic subperiod may be unknown. This trait is coded as present whenever an Archaic occupation is indicated, regardless of whether a finer temporal assignment is also made.

**Inclusion criteria:** Described in text OR indicated in structured fields (e.g., period field lists "Archaic," "Early Archaic," "Middle Archaic," "Late Archaic"; artifact field lists Archaic diagnostic projectile point types).

**Exclusion criteria:** Not described in text AND not indicated in structured fields. Paleoindian, Woodland, Caddo, or other non-Archaic components should not be coded as present here.

**Typical exemplars:** Period field states "Archaic." Narrative describes "a Late Archaic midden." Diagnostic projectile point types associated with the Archaic (e.g., Gary, Motley, Evans, Pontchartrain) are listed in the artifact inventory. Recorder notes "undifferentiated Archaic lithic scatter."

**Atypical exemplars:** No explicit period designation, but the artifact inventory lists only Gary or Evans points (well-established Archaic diagnostics in Louisiana) with no other temporal indicators. A site described as a shell midden that the recorder assigns to "prehistoric" but whose radiocarbon dates fall within the Archaic range.

**Close but no:** A site described as "prehistoric" with only generic lithic debitage and no diagnostic materials. Sites assigned to the Woodland period only. Sites where the recorder mentions "possible Archaic materials" without further support. Paleoindian components only.

**Response format:** Binary. Return `true` if the trait is present, `false` if absent.



## Chipped stone (Chipped\_Stone)

**Data type:** binary

**Short Description:** Presence of chipped stone artifacts, including tools and/or debitage.

**Definition:** The site form records the presence of any artifact produced primarily through knapping (flaking), including but not limited to: debitage (flakes, cores, angular shatter), bifaces, projectile points, scrapers, drills, perforators, and other flaked stone tools. Groundstone tools are excluded unless they also bear unambiguous knapping scars interpreted as intentional reduction.

**Inclusion criteria:** Described in text OR indicated in structured fields (e.g., artifact tally includes debitage, flakes, cores, bifaces, or other knapped tool categories; narrative mentions lithic scatter or chipped stone).

**Exclusion criteria:** Not described in text AND not indicated in structured fields. Sites with only groundstone, ceramics, or other non-knapped artifact classes should be coded as absent.

**Typical exemplars:** Artifact tally includes a count of "lithic debitage" or "flakes." Narrative states "chipped stone tools were recovered." Projectile points, scrapers, or bifaces are listed in the artifact inventory. The site is described as a "lithic scatter."

**Atypical exemplars:** No explicit mention of chipped stone, but a projectile point is listed in the artifact inventory (projectile points are chipped stone by definition). Narrative mentions "prehistoric artifacts" and the only illustrated or described items are flaked tools.

**Close but no:** Only groundstone (e.g., grinding stones, manos, metates) present, with no mention of knapped materials. Only ceramics and historic-era artifacts listed. The word "stone" appears only in the context of construction materials or geology with no artifact referent.

**Response format:** Binary. Return `true` if the trait is present, `false` if absent.



## Projectile point (Proj\_Point)

**Data type:** binary

**Short Description:** Presence of one or more projectile points.

**Definition:** The site form records the presence of at least one artifact identified as a projectile point — a bifacially or unifacially retouched tool interpreted as the tip or hafted element of a spear, atlatl dart, or arrow. Formal typological designations (e.g., "Gary point," "Alba point," "Perdiz point") are sufficient. Fragments described as probable projectile points are acceptable. Tools identified primarily as drills or perforators are excluded unless the recorder explicitly also identifies them as projectile points.

**Inclusion criteria:** Described in text OR indicated in structured fields (e.g., artifact tally includes "projectile point," "point," or a named type; narrative describes points or arrowheads).

**Exclusion criteria:** Not described in text AND not indicated in structured fields. General lithic debitage or undifferentiated flakes without mention of projectile points should be coded as absent.

**Typical exemplars:** Artifact tally lists "3 projectile points." Narrative states "a Gary point was recovered from the surface." The recorder lists "Alba points" or "arrowheads" in the artifact inventory. A point fragment is described as such.

**Atypical exemplars:** No explicit category labeled "projectile point," but the artifact inventory lists a named type that is unambiguously a projectile point form (e.g., "2 Perdiz," "1 Scallorn"). Narrative describes "a bifacial point tip fragment."

**Close but no:** Only debitage or cores listed with no mention of finished point forms. A "possible projectile point" noted by the recorder without further description. Drills or perforators listed but not described as projectile points. The word "point" appears only in a geographic context (e.g., "Poverty Point culture" or a place name). Descriptions of bifaces, biface preforms, or blanks without further specification of whether they represent projectile points.

**Response format:** Binary. Return `true` if the trait is present, `false` if absent.



## Fire-cracked rock (FCR)

**Data type:** binary

**Short Description:** Presence of fire-cracked rock (thermally fractured rock).

**Definition:** The site form records the presence of rock that has been fractured as a result of exposure to high heat, typically associated with cooking features, earth ovens, hearths, or burned rock middens. May be referred to as "fire-cracked rock," "FCR," "burned rock," "heat-fractured rock," "fire-altered rock," or similar terms. Rock that is merely discolored or stained without fracturing does not qualify.

**Inclusion criteria:** Described in text OR indicated in structured fields (e.g., artifact or feature tally includes FCR, burned rock, or a feature type associated with FCR such as "burned rock midden" or "earth oven").

**Exclusion criteria:** Not described in text AND not indicated in structured fields. Thermally discolored but non-fractured rock does not qualify. Burned structural materials (e.g., burned daub, burned brick) are excluded.

**Typical exemplars:** Artifact tally includes a count or weight of "FCR" or "fire-cracked rock." Narrative states "burned rock was recovered from a hearth feature." A feature is described as a "burned rock midden." The recorder notes "considerable fire-cracked limestone."

**Atypical exemplars:** The site is described as containing hearth features and the artifact inventory includes "burned material" if it is clear from context that this refers to rock rather than organic material or ceramics, code as present. A burned rock cluster is described in the feature section but not explicitly tallied in the artifact section.

**Close but no:** Only burned bone, burned ceramics, or burned organic material mentioned, with no reference to burned or fractured rock. Scorched sediment or ash described without associated fractured rock. A hearth feature is listed but no mention is made of associated rock. "Burned" applied only to structural daub or construction materials.

**Response format:** Binary. Return `true` if the trait is present, `false` if absent.



## NRHP Eligibility

**Data type:** categorical

**Short Description:** Status of site as a potential candidate for inclusion in the National Register of Historic places or as a protected resource. 

**Definition:** Eligibility statement outlining the current status of the site, or the recommendation for the site.

**Inclusion criteria:** Code if National Register eligibilty, site eligibility, research potential of the site as it relates to NRHP eligibility, or presence of graves or other protected resources is mentioned. 

**Exclusion criteria:** No discussion about national register eligibility, site eligibility, research potential, or protected status of the site. 

**Categories:** 

Eligible: The site is judged as eligible for inclusion. 

Listed: The site had been listed in the register.  

Potential/undetermined eligibility: The site is marked as having potential for inclusion in the register, or there is a note that more research is needed to determine eligibility status. This typically is paired with a recommendation for further study.

Ineligible: The site is marked as not being eligible for inclusion

Protected: The site is noted as not being eligible for inclusion, but is otherwise a protected resource, like a cemetery or homestead with graves.

Unknown: The particular site form, or record of study in a particular field campaign, has no information about recommendations for NRHP eligibility. 

**Response format:** Categorical/character entry containing one of the following: "Eligible", "Listed", "Potential/undetermined eligibility", "Ineligible", "Protected", "Unknown". 




## Site depth

**Data type:** numeric/character

**Short Description:** The depth of the deepest sediments at a site that contain archaeological material. 

**Definition:** The field identified depth of archaeological material at a site, typically identified through either shovel testing or excavation of 1x1m units.

**Inclusion criteria:** Descriptions of excavation results are present in structured fields or in narrative text, and those descriptions include discussion of buried sediments that contain archaeological artifacts or features. 

**Exclusion criteria:** Site is an unexcavated site that has largely architectural or surface features.

**Response format:** Return a plain number in cm below surface (e.g., `40` not `"40 cmbs"` or `"0.4 m"`). Convert to cm if reported in other units. If site was excavated, but there is not enough information to determine the depth of archaeological deposits, return "not reported". If no excavation occurred, return `"unknown"`.




## Site area

**Data type:** numeric/character

**Short Description:** The reported area of the archaeological site. 

**Definition:** The field identified area of the site, either in terms of two dimensions (e.g. the length of the site on the East West axis, and the length of the site on the North South axis), or as an areal measurement. 

**Inclusion criteria:** The area of the site is described either in narrative, or standardized fields. 

**Exclusion criteria:** The site is a landscape scale feature, network of features (like historic roads or trails), or otherwise linear features that have no clearly reported area. 

**Response format:** Return a number in square meters but do not include the unit (e.g. return: 450. Do not return: 450m^2. Convert to square meters if area is reported in other units, like square kilometers, acres, or hectares. If only dimensions are reported, calculate area from those dimensions assuming a rectlinear shape.






## Methodology

**Data type:** list

**Short Description:** Methodologies employed during the site visit reported in the form. 

**Definition:** Any one or more methodologies performed during the site visit. 

**Inclusion criteria:** Some kind of site visit, identification, or assessment is described.

**Exclusion criteria:** By definition, any of the site forms should have some form of methodology reported, at minimum the site was visited.

**Categories:** 

Surface survey: Site evaluate through surface survey

Shovel testing: Site evaluated through excavating one or more shovel test units

Test unit: Site evaluated through hand excavated test units, often in 1x1 meter cells

Backhoe trenching: Site evaluated through excavation of trenches by a backhoe or some other form of construction machinery

Architectural survey: Architectural assessment performed on features at the site

Historical records/interviews: Information about the site gathered through historical records or local informants

Geological: Formal geoarchaeological assessment


**Response format:** JSON array containing zero or more of: "Surface survey", "Shovel testing", "Test unit", "Backhoe trenching", "Architectural survey", "Historical records/interviews", "Geological"




## Tramline or rail infrastructure

**Data type:** binary

**Short Description:** Remains of rail-based transport, including logging tram grades and railroad earthworks.

**Definition:** The form identifies remains of a rail or tram corridor — a tram grade or railroad berm, flanking borrow ditches, cuts through rising ground.

**Inclusion criteria:** Described in text OR indicated in structured fields. Code present when the recorder names a tram, tramline, railroad, rail bed, rail grade, or spur, or describes a linear feature and argues that it served rail or tram transport. 

**Exclusion criteria:** Not described in text AND not indicated in structured fields. A linear raised feature, berm, or parallel ditch system recorded without the recorder attributing it to rail or tram use. Linear features the form attributes to another function — a levee, a drainage or irrigation ditch, a road, a logging skid trail, a pipeline or utility corridor, a firebreak.

**Typical exemplars:** "This site includes a tramline feature running NE/SW for 30 meters" 

**Atypical exemplars:**  "The site is disturbed by a historic tramline which cut the site at the NE corner"

**Close but no:** Berms and grades not described as relating to a tramline.

**Response format:** Binary. Return `true` if the trait is present, `false` if absent.



## Other logging infrastructure

**Data type:** binary

**Short Description:** archaeological evidence for substantial logging infrastructure independent of tramlines: mills, and related equipment.

**Definition:** The form describes archaeological remains of commercial timber extraction or processing that are not tramline or rail features. This covers remains of sawmill complexes and their footings, boilers, mill ponds, splash dams, and sawdust or slab deposits;  skidder sets, loading decks, and log landings; and the discarded hardware of those operations, such as cable, steam-donkey parts.

**Inclusion criteria:** Described in text OR indicated in structured fields. Code present when authors argue that the findings relate to a logging related installation or feature relating to the logging industry. 

**Exclusion criteria:** Not described in text AND not indicated in structured fields. Tramline and rail features, which are coded under Tramline or rail infrastructure. Logging camps and workers' quarters, including can scatters and levelled benches the recorder reads as a logging camp: this trait covers centralized installations and their equipment, not the encampments of the workforce. Individual artifacts, like saw fragments, where no installation or feature is described.

**Typical exemplars:** Concrete sawmill footings with an adjacent mill pond and slab pile. A sawdust or slab deposit the recorder attributes to a mill. A log landing or loading deck the recorder identifies as such.

**Atypical exemplars:** Structural piers with no other mill evidence described, where the recorder argues for a mill or related installation. Steam-donkey or skidder cable and parts the recorder reads as a skidder set.

**Close but no:** A tram berm with no other logging evidence described. Description of impacts to sites due to modern logging activities. Logging camps and workers' quarters. Can scatters described as relating to a logging camp.

**Response format:** Binary. Return `true` if the trait is present, `false` if absent.



## Turpentine cups

**Data type:** binary

**Short Description:** Clay or metal cups used to collect resin from catfaced pines.

**Definition:** The form describes turpentine cups — the Herty-system terracotta cups introduced in the early twentieth century, or the later metal cups used to catch resin running from a catface on a worked pine. Tin gutters or aprons nailed to the tree to direct resin into the cup fall under this trait.

**Inclusion criteria:** Described in text OR indicated in structured fields. Code present when the recorder records turpentine cups, Herty cups, or resin cups by any of those names, or the gutters and aprons associated with them. Fragments the recorder identifies as cup sherds are sufficient.

**Exclusion criteria:** Not described in text AND not indicated in structured fields. Undifferentiated historic ceramic or metal containers the recorder does not identify as turpentine-related. Catfaces or boxes with no cup described. Those are coded under "Other turpentine evidence".

**Typical exemplars:** "Several Herty cups were observed at the bases of catfaced pines." Terracotta turpentine cup fragments in a surface scatter. Metal resin cups and tin gutters collected.

**Atypical exemplars:** A scatter of coarse terracotta sherds the recorder identifies as cup fragments.

**Close but no:** Flower pot sherds. Undiagnostic coarse earthenware..

**Response format:** Binary. Return `true` if the trait is present, `false` if absent.




## Tar kiln

**Data type:** binary

**Short Description:** Potential earth kiln for rendering tar and pitch from fat lightwood.

**Definition:** The form describes a tar kiln: a broad, shallow, roughly circular depression, commonly 5 to 15 m across, where resinous lightwood was slow-burned under an earth cover to render tar and pitch. Charcoal, ash, or burned sand is frequently noted, and sometimes a prepared clay floor or a central collection channel running to a catchment at the edge. Related terms include pitch kiln, tar pit, and tar kiln scar. 

**Inclusion criteria:** Described in text OR indicated in structured fields. Code present when the recorder names a tar kiln, pitch kiln, or tar pit.

**Exclusion criteria:** Not described in text AND not indicated in structured fields. Pits the recorder attributes to gum turpentine production. Depressions the recorder attributes to another cause — a cellar, privy, well, borrow pit, treefall, or stock pond — or records without interpreting.

**Typical exemplars:** "A tar kiln approximately 8 m in diameter with a charcoal-stained floor is present on the ridge nose." The recorder maps and names a pitch kiln.

**Atypical exemplars:** A broad shallow circular depression with charcoal flecking that the recorder interprets as probable tar production without naming a kiln type.

**Close but no:** A rectangular depression interpreted as a structural cellar. A deep narrow pit. A modern burn pile. A turpentine still firebox. A broad shallow depression the recorder records but does not interpret. Such depressions are more appropriately coded as "Unattributed broad pit".

**Response format:** Binary. Return `true` if the trait is present, `false` if absent.



## Unattributed broad pit

**Data type:** binary

**Short Description:** A broad shallow depression of tar-kiln form that the recorder describes but does not interpret.

**Definition:** The form describes a pit or depression whose morphology is consistent with a tar kiln, broad, shallow, roughly circular, on the order of several metres across, but assigns it no function, or records its function as unknown. 
A broad shallow depression may be a tar kiln, a borrow pit, a collapsed cellar, a treefall, or a natural hollow.

**Inclusion criteria:** Described in text OR indicated in structured fields. Code present when the form describes a broad, shallow, roughly circular depression and either assigns it no function or records the function as unknown, uncertain, or undetermined. Measurements help but are not required where the description conveys the form.

**Exclusion criteria:** Not described in text AND not indicated in structured fields. Any depression the recorder identifies as a tar kiln, pitch kiln, or tar pit — those belong to Tar kiln. Any depression the recorder confidently attributes to another function: a structural cellar, privy, well, borrow pit, stock pond, treefall, or crater. Narrow or deep pits, rectangular depressions, and linear ditches, whose form is not that of a tar kiln.

**Typical exemplars:** "A shallow circular depression approximately 7 m in diameter is present on the ridge; function unknown." "Feature 2 is a broad basin-shaped depression of undetermined origin."

**Atypical exemplars:** A depression recorded only on the sketch map with a dimension and no label, where the narrative does not discuss it. A recorder noting "several shallow depressions of uncertain origin" without describing them individually.

**Close but no:** A depression the recorder calls a probable borrow pit. A cellar hole. A treefall the recorder identifies as such. A deep narrow shaft. A depression whose dimensions are not given and whose description does not convey a broad, shallow form.

**Response format:** Binary. Return `true` if the trait is present, `false` if absent.



## Other turpentine evidence

**Data type:** binary

**Short Description:** Naval-stores evidence that is neither a cup nor a pit.

**Definition:** The form describes evidence of turpentining or naval-stores production other than turpentine cups and tar kilns. This includes catface scars and chipped boxes on living pines or stumps; still and distillery remains; cooperage and barrel hoops; the hand tools of the trade, such as hacks, pullers, scrape irons, and dip buckets; and camps or quarters the recorder attributes to turpentine operations.

**Inclusion criteria:** Described in text OR indicated in structured fields. Code present when the recorder records naval-stores evidence falling outside the cup and pit traits, or argues that a feature or artifact relates to turpentine production.

**Exclusion criteria:** Not described in text AND not indicated in structured fields. Evidence already captured by Turpentine cups or Tar kiln. Historic material the recorder does not connect to naval stores.

**Typical exemplars:** "Numerous catfaced longleaf pines are present across the site." Remains of a turpentine still. Barrel hoops and a scrape iron the recorder attributes to turpentining. A quarters site the recorder assigns to a turpentine camp.

**Atypical exemplars:** A stand of stumps bearing paired diagonal scars that the recorder interprets as boxing, without using the word turpentine.

**Close but no:** Fire-scarred pines the recorder attributes to burning. Undiagnostic barrel hardware with no stated naval-stores association. A tar kiln, which is coded under Tar kiln.

**Response format:** Binary. Return `true` if the trait is present, `false` if absent.



## Homestead or farmstead

**Data type:** binary

**Short Description:** Remains of a rural domestic occupation, with or without associated agricultural features.

**Definition:** The form describes the remains of a dwelling and its associated yard and outbuildings. The form may also simply describe scatters of artifacts, and the author may intuit that it is a farmstead/homestead based on positioning on the landscape and other context clues. 

**Inclusion criteria:** Described in text OR indicated in structured fields. Code present when the form describes domestic structural remains, or a domestic refuse scatter in direct association with a structure, well, or cistern. Also code present where the recorder argues for a house site. Terms such as "old home place", "homestead", or "farmstead" are sufficient on their own.

**Exclusion criteria:** Not described in text AND not indicated in structured fields. A refuse dump with no structural or water-feature association and no argument by the recorder for a dwelling. Worker housing within a logging camp, unless the form describes independent domestic or agricultural occupation.

**Typical exemplars:** A chimney fall with surrounding whiteware, bottle glass, and cut nails. Brick piers, a hand-dug well, and a privy depression. "The site is the location of the old Dowden home place." A barn foundation with associated fencing.

**Atypical exemplars:** A levelled pad on a low rise with a bounded concentration of whiteware, bottle glass, and nails, which the recorder interprets as a house site despite no surviving foundation. A domestic refuse scatter with an associated hand-dug well and no other structural evidence.

**Close but no:** An isolated bottle dump in a borrow pit. A logging camp described purely as temporary quarters. A standing modern residence recorded as a land-use condition rather than as an archaeological site.

**Response format:** Binary. Return `true` if the trait is present, `false` if absent.


